Data_extractions_solutions

### 2019-02-02 Data_extractions.Rmd
Incorporated comments from Alfredo and Nick

Imagine, you quickly got well and instead of reproducing decided to carry on the Terminal Investment Meta-analysis Project.   
After a few weeks you fished searching and screening literature for the relevant papers. You now have a collection of around 30 included full-text papers (it could be 300 as well), stored neatly as pdf files and supplementary information, in whatever format available. Its time to start extracting the data (ideally, you should have planned this in detail while writing up your protocol/preregistration, but in this exercise we assume you did not).

@@

**QUESTION 4:**   
Whats wrong with this table representing a fragment of an extraction spreadsheet?

@@

**QUESTION 9:**   
What is meta-data in meta-analysis and what it is used for?

**Note:** In your actuall extraction sheet/form, try to group related columns together. You can also color-code them in Excel or put on the same tab/section in a visual database interface, if used. Your extraction form (spreadsheet) should be tested on several papers (pilot extractions) to make sure its easy to use and captures necessary information.

@@

**TASK 1**:  
We will practice data extraction on few of the Terminal investment papers. We will use R package *metaDigitise*. 
Detailed package description and tutorial are available at: https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html
https://github.com/daniel1noble/metaDigitise and this publication:
https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13118

Here is the package installation code:

@@

```{r load packages, eval=TRUE, message = FALSE}
library(metaDigitise)
library(tidyverse)
```

@@

You will need to open the pdf files and take the snapshots of the relevant figures. Make sure you capture the axes and axes labels. Also, enlarge these figures to fill the screen before taking the snapshot to get higher resolution image.

@@

**TASK 3**:  
We will now use *escalc* function from *metafor* package to calculate effect sizes. The function allows to calculate various effect sizes or outcome measures (and the corresponding sampling variances) that are commonly used in meta-analyses, including two effect sizes are relevant to our data and question:   
 * "SMD" stands for standarised mean difference, i.e. Hedges' g (potentially also "SMDH" for heteroscedastic variances).   
 * "ROM" for the log transformed ratio of means (Hedges et al., 1999; Lajeunesse, 2011), also named lnRR.   
We calculate them as follows:

@@

Open the above webpage. Main menu has an expandable list of types of effect sizes, and statistical tests you can use to calculate these effect sizes (only d, r, OR and RR).

**TASK 4**:   
Weil 2006 reports in the results: "At the end of the experiment saline-injected animals had smaller testes (F 1,27 = 4.6875, p<0.05, figure 2b)" and in the methods it is noted these are calculated using ANOVA.
Use online ES calculator sheet selecting "Standarized Mean difference (d)"" and "F-test, 2 groups" for to calculate d from one-way ANOVA (you will need F-value and sample sizes of treatment and control groups).

You should get d for the testes  mass that is close to what we calculated for tests volume from Fig2a. Note that this web calculator provides confidence intervals (CI) and variance for the calculated effect sizes.

**TASK 5**:  
Using web-based ES calculator, estimate a few other effect sizes using test (inferential) statistics reported in Sadd2006.

@@

**QUESTION 16:**    
Is the sign of effect size important? Imagine we use d, which represents mean_treatment - mean_control group trait values. How we would interpret d = 0.5 and how we would interpret d = -0.5?

@@

**QUESTION 19:**   
Although not mentioned in the original PRISMA checklist, it is important to share the collected data, meta-data and analysis code? Think of east at least 3 reasons for doing so.

**ANSWER 19:**   
Generally speaking, it is for transparency, reproducibility and giving back to the community.

>>>>

Imagine, you quickly got well and instead of reproducing decided to carry on the Terminal Investment Meta-analysis Project.   
After a few weeks you finished searching and screening literature for the relevant papers. You now have a collection of around 30 included full-text papers (it could be 300 as well), stored neatly as pdf files and supplementary information, in whatever format available. It is time to start extracting the data (ideally, you should have planned this in detail while writing up your protocol/preregistration, but in this exercise we assume you did not).

@@

**QUESTION 4:**   
What is wrong with this table representing a fragment of an extraction spreadsheet?

@@

**QUESTION 9:**   
Why we should record meta-data when extracting fdata for meta-analysis?

**Note:** In your actual extraction sheet/form, try to group related columns together. You can also color-code them in Excel or put on the same tab/section in a visual database interface, if used. Your extraction form (spreadsheet) should be tested on several papers (pilot extractions) to make sure its easy to use and captures necessary information.

@@

**TASK 1**:  
We will practice data extraction on few of the Terminal investment papers. We will use R package *metaDigitise*.   
Detailed package description and tutorial are available at: https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html or

https://github.com/daniel1noble/metaDigitise and this publication:  
https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13118

Here is the package installation code (run if you have not installed it earlier):

@@

Load the packages we will need:   
```{r load packages, eval=TRUE, message = FALSE}
library(metaDigitise)
library(tidyverse)
library(compute.es)
```

@@

You will need to open the pdf files and take the snapshots of the relevant figures (for MacOS the easiest is via Command+Shift+4 and then select the screen fragment; for Windows see https://www.windowscentral.com/how-take-windows-10-screenshot-7-best-techniques). Make sure you capture the axes and axes labels. Also, enlarge these figures to fill the screen before taking the snapshot to get higher resolution image.

@@

**TASK 3**:  
We will now use *escalc* function from *metafor* package to calculate effect sizes. The function allows to calculate various effect sizes or outcome measures (and the corresponding sampling variances) that are commonly used in meta-analyses, including two effect sizes that are relevant to our data and question:   
 * "SMD" stands for standarised mean difference, i.e. Hedges' g (potentially also "SMDH" for heteroscedastic variances).   
 * "ROM" for the log transformed ratio of means (Hedges et al., 1999; Lajeunesse, 2011), also named lnRR.   
We calculate them as follows:

@@

Open the above webpage. Main menu has an expandable list of types of effect sizes, and statistical tests you can use to calculate these effect sizes (only *d*, *r*, *OR* and *RR*).

**TASK 4**:   
Weil 2006 reports in the results: "At the end of the experiment saline-injected animals had smaller testes (F 1,27 = 4.6875, p<0.05, figure 2b)" and in the methods it is noted these are calculated using ANOVA.
Use online ES calculator sheet selecting "Standarized Mean difference (d)" and "F-test, 2 groups" to calculate *d* from one-way ANOVA (you will need F-value and sample sizes of treatment and control groups).

You should get *d* for the testes  mass that is close to what we calculated for tests volume from Fig2a. Note that this web calculator provides confidence intervals (CI) and variance for the calculated effect sizes.

**TASK 5**:   
Using web-based effect size calculator, estimate a few other effect sizes using test (inferential) statistics reported in Sadd2006.

In practice, we would preer to do it in R, rather than in aan online up or any other non-scriptable software that cannot socument our calculations. Luckily, the R package *compute.es* allows not only compute various effect sizes from descriptive statistics, but also from inferential statistics. 
r function inputs and names are as follows:
 * ANCOVA F-test: a.fes()
 * ANCOVA means: a.mes()
 * ANCOVA means (pooled sd): a.mes2()
 * ANCOVA p-value: a.pes()
 * ANCOVA t-test: a.tes()
 * Chi-squared (1 df): chies()
 * Correlation: res()
 * d-statistic: des()
 * Failure group (binary): failes()
 * F-test: fes()
 * Log odds ratio: lores()
 * Means: mes()
 * Means (pooled sd): mes2()
 * Proportions (binary): propes()
 * p-value: pes()
 * t-test: tes()
To learn more see the package vignette or https://www.rdocumentation.org/packages/compute.es/versions/0.2-4/topics/compute.es-package

As an example, we will use F value from **TASK 4**, note that functions from the *compute.es* packages automatically calcualte all available effect sizes for the given imput and if you run it on a data frame it will generate a new data.frame with several effect size estimates. Try:

```{r use compute.es, eval=TRUE, echo = TRUE, message = FALSE}
#single calculation:
a.fes(f = 4.6875, n.1 = 28, n.2 = 10, R = 0, q = 0) #where we enter F-value, sample sizes of treatment and control group, covariate correlation and number of covariates (we have none in ANOVA))

#for vectors of values (lieke in a data frame):
results_ES <- a.fes(f = c(4.6875, 4.6875), n.1 = c(28, 10), n.2 = c(10, 10), R = c(0, 0), q = c(0, 0))
str(results_ES)
```

@@

**QUESTION 16:**    
Is the sign of effect size important? Imagine we use *d*, which represents mean_treatment - mean_control group trait values. How we would interpret *d* = 0.5 and how we would interpret *d* = -0.5?

@@

**QUESTION 19:**   
Although not mentioned in the original PRISMA checklist, it is important to share the collected data, meta-data and analysis code? Think of at least 3 reasons for doing so.

### 2019-02-02 Data_extractions_solutions.Rmd
Incorporated comments from Alfredo and Nick

**ANSWER 4:**    
 1. Non-informative column names (which group should be nr 1 ?)   
 2. Second study has non-numerical values in the columns that should be numerical (NA is good for missing values, but should be used consistently)    
 3. Species column is a mix of Latin and common names (should be all Latin)

**ANSWER 5:**    
CohortID is used to indicate that the same cohorts/group of animals, of different sex (F, M), were measured repeatedly at different ages (60 and 120). These are not independent data points, and this is captured by CohortID.

@@

**ANSWER 9:**    
Meta-data is a detailed description of all extracted variables/columns and all other variables/columns derived from them. It is a part of the documentation of your meta-analysis and it is necessary when you or somebody else checks the data or tries to re-analyse or re-use it. It is a good practice to start creating such documentation when you start extractions and coding and update it as you make changes and progress to analyses.

@@

**ANSWER 10:**    
Ideally, two people would be extracting the data in parallel 9independently, in-duplicate), to minimise errors. However it is often not practical. In such case one person could be extracting and another person checking the extractions (there are always mistakes!).

@@

**ANSWER 12:**    
There are several pieces of software that helps digitising various types of plots and extracting the data (you can find a list here: https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13118), so it is often possible.  However, sometimes plot resolution can be too low, or data points / lines are overlapping and obscuring some of the values. In such case, you can try to contact the authors for the actual numbers or raw data used to make the plot.

>>>>

**ANSWER 4:**    
 1. Non-informative column names (which group should be nr 1 ?)   
 2. Second study has non-numerical values in the columns that should be numerical (NA is good for missing values, but should be used consistently)    
 3. Species column is a mix of Latin and common names (should be all Latin)   
 4. There is an extra dot in " 6.4      .""

**ANSWER 5:**    
CohortID is used to indicate that the same cohorts/group of animals, of different sex (F, M), were measured repeatedly at different ages (60 and 120). These are not independent data points, and this is captured by CohortID. For different study question this could be an unique identifier of various study sites or populations that need tp be distinguished within and between studies.

@@

**ANSWER 9:**    
Meta-data is a detailed description of all extracted variables/columns and all other variables/columns derived from them. More specifically it describes how the data was coded (e.g. the measurement units, how conversions were made, any decisions and assumptions that had to be made in general for a given column of data"; e.g. whether sex was coded as "F" and "M" and "B", and what these values mean, jhow missing values were coded, and anything else that would allow you or somebody to extract and code the same data). Meta-data is an essential part of the documentation of your meta-analysis and it is extremely useful when you or somebody else checks the data or tries to re-analyse or re-use it. It is a good practice to start creating such documentation when you start extractions and coding and update it as you make changes and progress to analyses.

@@

**ANSWER 10:**    
Ideally, two people would be extracting the data in parallel (independently, in-duplicate), to minimise errors. However it is often not practical. In such case one person could be extracting and another person checking the extractions (there are always mistakes!).

@@

**ANSWER 12:**    
There are several pieces of software that helps digitising various types of plots and extracting the data (you can find a list here: https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13118, so it is often possible.  However, sometimes plot resolution can be too low, or data points / lines are overlapping and obscuring some of the values. In such case, you can try to contact the authors for the actual numbers or raw data used to make the plot.
